Optimizing the power of adaptive seamless designs using Bayesian Optimization

### 2020-12-09 main.tex
TODO notes

%\DOIsuffix{bimj.DOIsuffix}
\DOIsuffix{bimj.200100000}
\Volume{52}
\Issue{61}
\Year{2020}
\pagespan{1}{}
\keywords{Adaptive Seamless Designs; Clinical Trials; Model-Based Optimization; Treatment Selection\\\\ %FIXME: name 5 %Up to five keywords are allowed and should be given in alphabetical order. Please capitalize the key
%\noindent \hspace*{-4pc}\\
%\hspace*{-4pc} 
%{\small\it words)}\\[1pc]
\noindent\hspace*{-4.2pc} Supporting Information for this article is available from the author or on the WWW under\break \hspace*{-4pc} \underline{http://dx.doi.org/10.1022/bimj.XXXXXXX} (please delete if not
applicable)
}  %%% semicolon and fullpoint added here for keyword style

@@

% TODO: Jakob 
% phase 1 idea something fast
% phase 2
% Describe the study we look at. / The Data
%COPD trial

@@

\subsection{Two-stage design with treatment selection}
\label{ssec:two_stage_design_with_treatment_selection}
Here we consider a particular type of adaptive seamless designs, namely two-stage designs with treatment (or more specifically dose) selection. The trial starts in stage 1 with $K$ experimental treatments or doses of one experimental treatment which are compared to a common control. Interest is in assessing the treatment effects $\beta_k$ with $k=1,\dots, K$ comparing experimental treatment $k$ vs. control and testing the null hypotheses $H_{0,k}: \beta_k=0$ against the alternative  $H_{1,k}: \beta_k>0$ (larger $\beta_k$ being better). For the purpose of treatments selection, test statistics $Z_{1,k}$ based on data from the first design stage are considered with larger $Z_{1,k}$ indicating more beneficial treatments. These can use data on the primary outcome used for testing or any other outcome that is available at the time of the interim analysis. In practical applications, early outcomes are often used to inform interim decisions, since the primary endpoint might take a longer time to observe. In the interim analysis, treatments are selected for continuation into the second stage; the set of selected treatments is denoted by $\mathcal{S}$. The selection rules used here are described in Section \ref{ssec:selection_rules}. In the final analysis, the closed test principle is applied to account for the multiple hypotheses. The intersection hypotheses are tested by combining stagewise p-values for the respective intersection hypotheses using a prespecified combination function. Here we use the inverse normal combination function~\citep{lehmacher_adaptive_1999}  % Lehmacher and Wassmer https://doi.org/10.1111/j.0006-341X.1999.01286.x.
%FIXME Show function

@@

Figure~\ref{fig:plot_boxplot_valid_y} shows the estimated power values $y_{\text{valid}}$ for the best solutions in each scenario.
\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y.pdf}
\caption{%
  Box plots of performance (power) values on independent replicates of the best solutions for the four optimizers, the different effect sets, and different numbers of $n_{\text{treat}}$. %FIXME rename n_cases in plot
  }
\label{fig:plot_boxplot_valid_y}
\end{figure}

The plots show that \emph{MBO} generally gives results that are comparable to those obtained with the (exhaustive) \emph{Grid} method, for all effects sets. 
Only in some cases \emph{Grid} has a slight advantage (\emph{effect: linear}, $n_{\text{treat}} = 2000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$).
Below we will see that these scenarios have in common that the optimal configuration is close to the border of the search space. %FIXME Make sure that this is mentioned for figure 4!
In other cases \emph{MBO} yields slightly better results (\emph{effect: paper}, $n_{\text{treat}} = 2000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
Across all scenarios, \emph{MBO} and \emph{Grid} perform similar, even though \emph{MBO} due to its evaluation budget is allowed only 116 evaluations in contrast to the 1350 evaluations for \emph{Grid}, with a resolution of 25 points per dimension. 
The difference in the number of evaluations is clearly reflected in the average runtime of the optimization methods given in Table~\ref{tab:table_time}.
\emph{MBO} is approximately 20 times faster.
\input{generated/tables/table_time.tex}

@@

Comparing \emph{MBO Grid} to \emph{Gird} we can observe, that for the scenarios \emph{effect: paper}, $n_{\text{treat}} = 500$ and \emph{effect: paper}, $n_{\text{treat}} = 1000$ \emph{MBO Grid} achieves slightly better performance then \emph{Grid}.
The \emph{Grid} optimization selected the best outcome "optimistically" from all observed outcome and has no notion of the stochasticity of the problem, i.e. in this case the noise on the predicted power. 
Accordingly, we assume that for the mentioned scenarios the noise was so high, that \emph{Gird} was misguided by overoptimistic outcomes when selecting the final configuration as optimization result.
The independent validation $y_{\text{valid}}$ represents the unbiased performance of this final configuration, and the overoptimism becomes apparent.
In these scenarios \emph{MBO} performs better, since the final configuration is determined based on the mean prediction of the surrogate of \emph{MBO}, as explained in Section~\ref{ssec:best_point}.
Below we will study on a single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of \emph{Grid}. %FIXME: n_treat 500, paper 2 would be more interesting or n_treat 1000 and linear)

@@

Figure~\ref{fig:plot_opt_path_5000} shows that the lower noise level leads to more stable simulation runs, meaning that each single optimization run leads to a results closer to the mean result.
Furthermore, the average optimization curve indicates that less iterations can be sufficient to obtain an outcome that is comparable to \emph{Grid},
It has to be noted, that this plot reports the best observed outcome so far and not the one that is chosen as the final result.
So on the left side of Figure~\ref{fig:plot_opt_path_5000}, we see more curves that observed overoptimistic outcomes in comparison the the right side.
Ideally, the choice of the final point should not be influenced by these overoptimistic observations for \emph{MBO} as explained in Section~\ref{ssec:best_point}, as the final point is determined by the surrogates mean prediction.
This "bias correction" seems to be effective, as the validation error reported in Figure~\ref{fig:plot_boxplot_valid_y_5000} is nearly identical for \emph{MBO} for the low and the high value of $n_\text{sim}$.
%FIXME: Add Runtime comparision? Maybe boring because it probably is just 5 times the old values
\begin{figure}[htb]
\centering
  \begin{minipage}{0.62\textwidth}
    \centering
    \includegraphics[width=\linewidth]{generated/figures/plot_opt_path_5000.pdf}%
    \captionof{figure}{Optimization progress for \emph{MBO} in each stochastic repetition in comparison to the best found outcome in the grid search on effect set \emph{paper} and $n_{\text{treat}} = 2000$.}
    \label{fig:plot_opt_path_5000}
  \end{minipage}\hspace{0.02\textwidth}%
  \begin{minipage}{0.35\textwidth}
    \centering
    \includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y_5000.pdf}%
    \captionof{figure}{Validated performance of all optimization strategies on effect set \emph{paper} and $n_{\text{treat}} = 2000$ with 100 stochastic repetitions and different values for \texttt{nsim}}
    \label{fig:plot_boxplot_valid_y_5000}
  \end{minipage}
\end{figure}
For all methods the IQR displayed in the Box plots is much smaller for $n_\text{sim} = 5000$, indicating that the found solutions vary less within the 100 stochastic replications.
Also the absolute differences between the methods seem even more negligible.
This demonstrates, that for the selected scenario all methods can find the optimal configuration if the results are not noisy.
But as an increased value of $n_\text{sim}$ also increases the runtime linearly, increasing $n_\text{sim}$ is not always a feasible choice if the goal is to find the optimal configuration quickly.
We conclude with the observation, that \emph{MBO} can find configurations with a low number of needed simulations (higher noise) that are close to the optimal found solution with a high number of simulation.

>>>>

%\DOIsuffix{bimj.DOIsuffix}
\DOIsuffix{bimj.200100000}
\Volume{52}
\Issue{61}
\Year{2020}
\pagespan{1}{}
\keywords{Adaptive Seamless Designs; Bayesian Optimization; Clinical Trials; Treatment Selection\\\\ %Up to five keywords are allowed and should be given in alphabetical order. Please capitalize the key
%\noindent \hspace*{-4pc}\\
%\hspace*{-4pc} 
%{\small\it words)}\\[1pc]
\noindent\hspace*{-4.2pc} Supporting Information for this article is available from the author or on the WWW under\break \hspace*{-4pc} \underline{http://dx.doi.org/10.1022/bimj.XXXXXXX} (please delete if not
applicable)
}  %%% semicolon and fullpoint added here for keyword style

@@

@@

\subsection{Two-stage design with treatment selection}
\label{ssec:two_stage_design_with_treatment_selection}
Here we consider a particular type of adaptive seamless designs, namely two-stage designs with treatment (or more specifically dose) selection. The trial starts in stage 1 with $K$ experimental treatments or doses of one experimental treatment which are compared to a common control. Interest is in assessing the treatment effects $\beta_k$ with $k=1,\dots, K$ comparing experimental treatment $k$ vs. control and testing the null hypotheses $H_{0,k}: \beta_k=0$ against the alternative  $H_{1,k}: \beta_k>0$ (larger $\beta_k$ being better). For the purpose of treatments selection, test statistics $Z_{1,k}$ based on data from the first design stage are considered with larger $Z_{1,k}$ indicating more beneficial treatments. These can use data on the primary outcome used for testing or any other outcome that is available at the time of the interim analysis. In practical applications, early outcomes are often used to inform interim decisions, since the primary endpoint might take a longer time to observe. In the interim analysis, treatments are selected for continuation into the second stage; the set of selected treatments is denoted by $\mathcal{S}$. The selection rules used here are described in Section \ref{ssec:selection_rules}. In the final analysis, the closed test principle is applied to account for the multiple hypotheses. The intersection hypotheses are tested by combining stagewise p-values for the respective intersection hypotheses using a prespecified combination function. Here we use the inverse normal combination function~\citep{lehmacher_adaptive_1999}
%FIXME Show function

@@

Figure~\ref{fig:plot_boxplot_valid_y} shows the estimated power values $y_{\text{valid}}$ for the best solutions in each scenario.
\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y.pdf}
\caption{%
  Box plots of performance (power) values on independent replicates of the best solutions for the four optimizers, the different effect sets, and different numbers of $n_{\text{treat}}$.
  }
\label{fig:plot_boxplot_valid_y}
\end{figure}

The plots show that \emph{MBO} generally gives results that are comparable to those obtained with the (exhaustive) \emph{Grid} method, for all effects sets. 
Only in some cases \emph{Grid} has a slight advantage (\emph{effect: linear}, $n_{\text{treat}} = 2000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$).
Below we will see that these scenarios have in common that the optimal configuration is close to the border of the search space. 
In other cases \emph{MBO} yields slightly better results (\emph{effect: paper}, $n_{\text{treat}} = 2000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
Across all scenarios, \emph{MBO} and \emph{Grid} perform similar, even though \emph{MBO} due to its evaluation budget is allowed only 116 evaluations in contrast to the 1350 evaluations for \emph{Grid}, with a resolution of 25 points per dimension. 
The difference in the number of evaluations is clearly reflected in the average runtime of the optimization methods given in Table~\ref{tab:table_time}.
\emph{MBO} is approximately 20 times faster.
\input{generated/tables/table_time.tex}

@@

Comparing \emph{MBO Grid} to \emph{Gird} we can observe, that for the scenarios \emph{effect: paper}, $n_{\text{treat}} = 500$ and \emph{effect: paper}, $n_{\text{treat}} = 1000$ \emph{MBO Grid} achieves slightly better performance then \emph{Grid}.
The \emph{Grid} optimization selected the best outcome "optimistically" from all observed outcome and has no notion of the stochasticity of the problem, i.e. in this case the noise on the predicted power. 
Accordingly, we assume that for the mentioned scenarios the noise was so high, that \emph{Gird} was misguided by overoptimistic outcomes when selecting the final configuration as optimization result.
The independent validation $y_{\text{valid}}$ represents the unbiased performance of this final configuration, and the overoptimism becomes apparent.
In these scenarios \emph{MBO} performs better, since the final configuration is determined based on the mean prediction of the surrogate of \emph{MBO}, as explained in Section~\ref{ssec:best_point}.
Below we will study on a single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of \emph{Grid}.

@@

Figure~\ref{fig:plot_opt_path_5000} shows that the lower noise level leads to more stable simulation runs, meaning that each single optimization run leads to a results closer to the mean result.
Furthermore, the average optimization curve indicates that less iterations can be sufficient to obtain an outcome that is comparable to \emph{Grid},
It has to be noted, that this plot reports the best observed outcome so far and not the one that is chosen as the final result.
So on the left side of Figure~\ref{fig:plot_opt_path_5000}, we see more curves that observed overoptimistic outcomes in comparison the the right side.
Ideally, the choice of the final point should not be influenced by these overoptimistic observations for \emph{MBO} as explained in Section~\ref{ssec:best_point}, as the final point is determined by the surrogates mean prediction.
This "bias correction" seems to be effective, as the validation error reported in Figure~\ref{fig:plot_boxplot_valid_y_5000} is nearly identical for \emph{MBO} for the low and the high value of $n_\text{sim}$.
\begin{figure}[htb]
\centering
  \begin{minipage}{0.62\textwidth}
    \centering
    \includegraphics[width=\linewidth]{generated/figures/plot_opt_path_5000.pdf}%
    \captionof{figure}{Optimization progress for \emph{MBO} in each stochastic repetition in comparison to the best found outcome in the grid search on effect set \emph{paper} and $n_{\text{treat}} = 2000$.}
    \label{fig:plot_opt_path_5000}
  \end{minipage}\hspace{0.02\textwidth}%
  \begin{minipage}{0.35\textwidth}
    \centering
    \includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y_5000.pdf}%
    \captionof{figure}{Validated performance of all optimization strategies on effect set \emph{paper} and $n_{\text{treat}} = 2000$ with 100 stochastic repetitions and different values for \texttt{nsim}}
    \label{fig:plot_boxplot_valid_y_5000}
  \end{minipage}
\end{figure}
For all methods the IQR displayed in the Box plots is much smaller for $n_\text{sim} = 5000$, indicating that the found solutions vary less within the 100 stochastic replications.
Also the absolute differences between the methods seem even more negligible.
This demonstrates, that for the selected scenario all methods can find the optimal configuration if the results are not noisy.
But as an increased value of $n_\text{sim}$ also increases the runtime linearly, increasing $n_\text{sim}$ is not always a feasible choice if the goal is to find the optimal configuration quickly.
We conclude with the observation, that \emph{MBO} can find configurations with a low number of needed simulations (higher noise) that are close to the optimal found solution with a high number of simulation.
